---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = NA
)
```

# traveltime

<!-- badges: start -->
[![traveltime status badge](https://idem-lab.r-universe.dev/badges/traveltime)](https://idem-lab.r-universe.dev/traveltime)
[![R-CMD-check](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/idem-lab/traveltime/branch/main/graph/badge.svg)](https://app.codecov.io/gh/idem-lab/traveltime?branch=main)
![GitHub License](https://img.shields.io/github/license/geryan/traveltime)
[![Lifecycle:](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
![GitHub commit activity](https://img.shields.io/github/commit-activity/w/geryan/traveltime)
![GitHub Downloads (all assets, all releases)](https://img.shields.io/github/downloads/geryan/traveltime/total)
![GitHub last commit](https://img.shields.io/github/last-commit/geryan/traveltime)
![GitHub commits since latest release](https://img.shields.io/github/commits-since/geryan/traveltime/latest)
<!-- badges: end -->

Implements methods from Weiss et al. 2018, 2020 to calculate travel time from
given locations over a friction surface.

Citations:

*D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C. Kamath,
A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H. Keddie, S.
Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar, T. L. Symons,
E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething.* **Global maps of travel
time to healthcare facilities.** (2020) Nature Medicine.
https://doi.org/10.1038/s41591-020-1059-1

*D. J. Weiss, A. Nelson, H.S. Gibson, W. Temperley, S. Peedell, A. Lieber, M.
Hancher, E. Poyart, S. Belchior, N. Fullman, B. Mappin, U. Dalrymple, J. Rozier,
T.C.D. Lucas, R.E. Howes, L.S. Tusting, S.Y. Kang, E. Cameron, D. Bisanzio, K.E.
Battle, S. Bhatt, and P.W. Gething.* **A global map of travel time to cities to
assess inequalities in accessibility in 2015.** (2018). Nature.
doi:10.1038/nature25181.

## Installation

You can install `traveltime` with:

``` r
install.packages("traveltime", repos = c("https://idem-lab.r-universe.dev"))
```

## Let's calculate some travel times

First download a friction surface — here using the motorised travel time from
Weiss *et al.* 2020.
```{r }
library(traveltime)
library(terra)

friction_surface <- get_friction_surface(
    surface = "motor2020",
    extent = c(111,112,0,1)
  )
friction_surface
```

Let's have a look at that
```{r}
plot(friction_surface)
```

Prepare points we would like to calculate travel time from
```{r}
from_here <- tibble::tibble(
  x = c(111.2, 111.9),
  y = c(0.2, 0.35)
)
from_here
```

And calculate the travel time
```{r}
travel_time <- calculate_travel_time(
  friction_surface = friction_surface,
  points = from_here
)
travel_time
```

Et voila!
```{r}
plot(travel_time)
points(from_here, pch = 19)
```

## Let's go to Singapore

Here it is:
```{r}
library(sdmtools)
sin <- sdmtools::make_africa_mask(
  type = "vector",
  countries = "SGP"
)
plot(sin)
```

We're going to see how long it takes to walk home from Changi Airport. So we'll 
download the walking-only friction surface this time.

We can feed in our `sin` `SpatVector` directly as the `extent`, instead of specifying by hand as above. 
```{r }
library(traveltime)
library(terra)

friction_singapore <- get_friction_surface(
    surface = "walk2020",
    extent = sin
  )|> 
  mask(sin)

friction_singapore
```

And where is Changi Airport?
```{r}
changi_airport <- tibble::tibble(
  x = c(103.984),
  y = c(1.355)
)
changi_airport
```
Let's look at those.
```{r}
plot(friction_singapore)
plot(sin, add = TRUE)
points(changi_airport, pch = 19)
```

And calculate the travel time
```{r}
travel_time_sin <- calculate_travel_time(
  friction_surface = friction_singapore,
  points = changi_airport
)
travel_time_sin
```

Et voi*lah*!
```{r}
plot(travel_time_sin)
points(changi_airport, pch = 19)
plot(sin, add = TRUE)
```


### I want my plots to be nicer
```{r}
library(tidyterra)
library(ggplot2)

ggplot() +
  # plot the spatraster first
  geom_spatraster(
    data = travel_time_sin
  ) +
  theme_minimal() +
  scale_fill_whitebox_c(palette = "deep") +
  # overlay the vector outline
  geom_spatvector(
    data = sin,
    colour = "grey70",
    linewidth = 1,
    fill = NA
  ) +
  # add the points from tibble
   geom_point(
     data = changi_airport,
     aes(x = x, y = y),
     colour = "hotpink"
   ) +
  labs(x = NULL, y = NULL, fill = "Travel time\n(minutes)") 

```


